Radiance - A Web Application Environment

### 2016-12-17 els2017-radiance/paper.tex
Fixed "This .." without subject.

%% FIXME:
% Use of "this" without being followed by a noun. Eg "this environment sharing", "this kind of repetition"
% Tweak minted colours

@@

\begin{abstract}
  Radiance\cite{radiance} is a set of libraries that provides an environment for web applications. Unlike traditional web frameworks that focus on a set of tools to support the construction and maintenance of a single application, Radiance attempts to allow one to run as many differing applications as desired in the same environment. This has many advantages, as frequently there are common resources in applications that should be shared between them, but cannot, because the applications are separated and developed without regard for sharing. In order to fix this, Radiance provides an extensible interface mechanism to standardise the interaction between users of common resources. To facilitate the separation and encapsulation of the shared URL address space between applications, Radiance also provides a flexible routing mechanism. This mechanism is powerful enough to ensure that any application written with Radiance can be deployed on any server setup, no matter how complicated the constraints, without needing to change the source code of the application.
\end{abstract}

@@

\section{Introduction}
Writing a website or web application is usually not a very complex or complicated process; doing so involves performing very similar tasks every time. This lead to the rise of web frameworks-- collections of tools and libraries to take care of the most common of aspects for the developer. \\

@@

Often these frameworks do not address many questions about deployment and server setup. The assumption is made that the application will be deployed in such a fashion that it is run on a single public-facing domain, which it has complete control over. More specifically, it is assumed that the full path of the URL can be controlled. This leads to complications when multiple applications should be run on the same server. Sharing can be accomplished by using different ports for each application's server, and then using a proxying server that defers requests to the appropriate application depending on the subdomain of the request. \\

@@

This leads to a divide in application design. If the task is to write an application that others should be able to use on their own setup, then it needs to be written in a particular way in order to account for the aforementioned setup differences. The framework will usually not help with this in any way, and an ad-hoc solution has to be written. As an example, Wordpress requires the configuration of a special variable to account for this.\cite{wordpress} \\

For administrators the problems that stem from these single-application focused frameworks do not end there, however. Many applications will need access to the same kinds of resources. Frequently, information about users, authorisation, sessions, and so forth should be shared, but cannot, because the various applications were designed individually, without any concern for standardised access of such resources. As such, running multiple applications on the same server results in duplication of information. This can be especially cumbersome for the users, as they need to track multiple registrations, profiles, and so forth, despite being on the same service. \\

@@

\subsection{Interfaces}
The first piece that came out of this line of thinking is an interface system. This system allows the specification of signatures of a package. This means that functions, macros, variables, etc. are defined in their signature only, without any actual functionality being implemented that makes them work. For example, a simplified interface to a database might look like this:

@@

Radiance provides standard interface definitions for most of the common aspects of web application programming. This allows the sharing of the previously discussed common resources. However, it is also sensible to define interfaces for commonly needed tools that may have slightly different implementation details. An example for this would be a caching system; the exact behaviour of the cache can vary --be it database-backed, file-backed, memory-backed, etc.-- but the usage of the system is the same, regardless of implementation details. \\

@@

Since the implementation mechanism does not rely on some sort of indirection facility, interface functionality incurs zero overhead, allowing implementations to be as efficient as possible. This is one of the details of the system that are rather specific to Lisp, since it makes this kind of behaviour trivial to implement. \\

Thus, the interface mechanism solves the problem of shared access across applications without constraining the system to a particular implementation. This allows the administrator of the system to choose the appropriate implementation for his setup, without having to take care to respect the wishes of every application. \\

@@

The routing system lies between the HTTP server and the web applications. This leads to the following request lifecycle for a typical Radiance setup: \\

@@

As it stands, Radiance would translate a request to \\\texttt{http://radiance.example.com/test/index} into the URI \\\texttt{radiance.example.com/test/index} and then map it to \\\texttt{radiance/test/index}. The top-level domain --the part of the domain we cannot change-- is stripped away by one of the standard routes. This is still not a URI that would invoke our page though. Therefore, additional routes need to be set up to handle the conversion to and from the internal representation.

@@

While the above routes are fairly simplistic in their behaviour, routes can be written to perform arbitrary operations, including stashing away data for later use during reversal. This allows things like the virtual subdomain route that is included by default. This route recognises when a request like this \texttt{/!/subdomain/foo} and maps it to \texttt{subdomain/foo}. It is necessary for the corresponding reversal route to know that the initial request was mapped by this particular route in order to appropriately translate back to the prefix, as there is no indication in the URI that we want to reverse that this should happen. \\

In order for everything to fit together, applications need to abide by some conventions that Radiance postulates. The most important one is that each application gets its own subdomain in the internal representation. This allows applications to assume that they have free reign over the path part of the URL. Additionally, the path prefixes \texttt{/static/} and \texttt{/api/} are reserved by Radiance on all domains, and are used to provide static files and REST api endpoints respectively. This way, routes can be written in a more generic manner to solve common problems, rather than requiring specifically tailored rules for each application. \\

@@

\subsection{Resources}
The final piece is a generic resource mechanism that allows the exchange of information between applications, and particularly between applications and interface implementations. This is necessary because applications may need to refer to resources provided by other components in the Radiance environment. For example, the login page and mechanism is provided by the authentication interface. An application may want to present a link to this page to the users. However, because the location of this page is not standardised, the application has no direct way of knowing where the page would be. Instead of specifying the page location, Radiance provides a system that allows the retrieval of such resources generically. \\

>>>>

%% FIXME:
% Tweak minted colours
% Clarify different perspective (admin vs dev)

@@

\begin{abstract}
  Radiance\cite{radiance} is a set of libraries that provides an environment for web applications. Unlike traditional web frameworks that focus on a set of tools to support the construction and maintenance of a single application, Radiance attempts to allow one to run as many differing applications as desired in the same environment. This goal brings many advantages, as frequently there are common resources in applications that should be shared between them, but cannot, because the applications are separated and developed without regard for sharing. In order to fix this, Radiance provides an extensible interface mechanism to standardise the interaction between users of common resources. To facilitate the separation and encapsulation of the shared URL address space between applications, Radiance also provides a flexible routing mechanism. This mechanism is powerful enough to ensure that any application written with Radiance can be deployed on any server setup, no matter how complicated the constraints, without needing to change the source code of the application.
\end{abstract}

@@

\section{Introduction}
Writing a website or web application is usually not a very complex or complicated process; doing so involves performing very similar tasks every time. This trend lead to the rise of web frameworks-- collections of tools and libraries to take care of the most common of aspects for the developer. \\

@@

Often these frameworks do not address many questions about deployment and server setup. The assumption is made that the application will be deployed in such a fashion that it is run on a single public-facing domain, which it has complete control over. More specifically, it is assumed that the full path of the URL can be controlled. This assumption leads to complications when multiple applications should be run on the same server. Sharing can be accomplished by using different ports for each application's server, and then using a proxying server that defers requests to the appropriate application depending on the subdomain of the request. \\

@@

This problem leads to a divide in application design. If the task is to write an application that others should be able to use on their own setup, then it needs to be written in a particular way in order to account for the aforementioned setup differences. The framework will usually not help with this in any way, and an ad-hoc solution has to be written. As an example, Wordpress requires the configuration of a special variable to account for this.\cite{wordpress} \\

For administrators the problems that stem from these single-application focused frameworks do not end there, however. Many applications will need access to the same kinds of resources. Frequently, information about users, authorisation, sessions, and so forth should be shared, but cannot, because the various applications were designed individually, without any concern for standardised access of such resources. As such, running multiple applications on the same server results in duplication of information. This duplication can be especially cumbersome for the users, as they need to track multiple registrations, profiles, and so forth, despite being on the same service. \\

@@

\subsection{Interfaces}
The first piece that came out of this line of thinking is an interface system. This system allows the specification of signatures of a package. What this means is that functions, macros, variables, etc. are defined in their signature only, without any actual functionality being implemented that makes them work. For example, a simplified interface to a database might look like this:

@@

Radiance provides standard interface definitions for most of the common aspects of web application programming, allowing the sharing of the previously discussed common resources. However, it is also sensible to define interfaces for commonly needed tools that may have slightly different implementation details. An example for this would be a caching system; the exact behaviour of the cache can vary --be it database-backed, file-backed, memory-backed, etc.-- but the usage of the system is the same, regardless of implementation details. \\

@@

Since the implementation mechanism does not rely on some sort of indirection facility, interface functionality incurs zero overhead, allowing implementations to be as efficient as possible. This advantage is one of the details of the system that are rather specific to Lisp, since it makes this kind of behaviour trivial to implement. \\

Thus, the interface mechanism solves the problem of shared access across applications without constraining the system to a particular implementation. Thanks to this, the administrator of the system can choose the appropriate implementation for his setup, without having to take care to respect the wishes of every application. \\

@@

The routing system lies between the HTTP server and the web applications, leading to the following request lifecycle for a typical Radiance setup: \\

@@

As it stands, Radiance would translate a request to \\\texttt{http://radiance.example.com/test/index} into the URI \\\texttt{radiance.example.com/test/index} and then map it to \\\texttt{radiance/test/index}. The top-level domain --the part of the domain we cannot change-- is stripped away by one of the standard routes. This URI is still not one that would invoke our page though. Therefore, additional routes need to be set up to handle the conversion to and from the internal representation.

@@

While the above routes are fairly simplistic in their behaviour, routes can be written to perform arbitrary operations, including stashing away data for later use during reversal. This flexibility allows things like the virtual subdomain route that is included by default. This route recognises when a request like this \texttt{/!/subdomain/foo} and maps it to \texttt{subdomain/foo}. It is necessary for the corresponding reversal route to know that the initial request was mapped by this particular route in order to appropriately translate back to the prefix, as there is no indication in the URI that we want to reverse that this should happen. \\

In order for everything to fit together, applications need to abide by some conventions that Radiance postulates. The most important one is that each application gets its own subdomain in the internal representation. This convention allows applications to assume that they have free reign over the path part of the URL. Additionally, the path prefixes \texttt{/static/} and \texttt{/api/} are reserved by Radiance on all domains, and are used to provide static files and REST api endpoints respectively. This way, routes can be written in a more generic manner to solve common problems, rather than requiring specifically tailored rules for each application. \\

@@

\subsection{Resources}
The final piece is a generic resource mechanism that allows the exchange of information between applications, and particularly between applications and interface implementations. This piece is necessary because applications may need to refer to resources provided by other components in the Radiance environment. For example, the login page and mechanism is provided by the authentication interface. An application may want to present a link to this page to the users. However, because the location of this page is not standardised, the application has no direct way of knowing where the page would be. Instead of specifying the page location, Radiance provides a system that allows the retrieval of such resources generically. \\
